Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit

### 2024-04-19 optim.qmd
outline of the section visualisation of jellyfish optimiser

```{=html}
<!--
There exist R packages to fit the hyperparameters of a Gaussian process, for example ExaGeoStatR (https://github.com/ecrc/exageostatR), which is an R wrapper for the ExaGeoStat library (https://arxiv.org/pdf/1708.02835.pdf).
-->
```

How can we define a measure of smoothness from this? The distribution of
a Gaussian process is fully determined by its mean function and
covariance function. The way the covariance function is defined is where
smoothness comes into play: if an index is very smooth, then two close
projection angles should produce close index values (strong
correlation); by contrast, if an index is not smooth, then two close
projection angles might give very different index values (fast decay of
correlations with respect to distance between angles).

@@

>>>>

```{=html}
<!--
There exist R packages to fit the hyperparameters of a Gaussian process, for example ExaGeoStatR (https://github.com/ecrc/exageostatR), which is an R wrapper for the ExaGeoStat library (https://arxiv.org/pdf/1708.02835.pdf).
-->
```
How can we define a measure of smoothness from this? The distribution of
a Gaussian process is fully determined by its mean function and
covariance function. The way the covariance function is defined is where
smoothness comes into play: if an index is very smooth, then two close
projection angles should produce close index values (strong
correlation); by contrast, if an index is not smooth, then two close
projection angles might give very different index values (fast decay of
correlations with respect to distance between angles).

@@

# Visualisation of jellyfish optimiser

Information of the jellyfish optimiser is available in a tabular format
and below is an example data collected from finding the sine-wave
structure in 6D data using a distance correlation index (`docr2d_2`):

```{r}
load(here::here("data/sim_sine_6d_dcor2d.rda"))
```

```{r}
dplyr::glimpse(sim_sine_6d_dcor2d |> select(-id)) 
```

Information recorded can be categorised into the following categories:

-   projection pursuit variables: the index function used (`idx_f`), the
    data dimension (`d`)
-   jellyfish optimiser parameters: the number of jellies (`n_jellies`),
    the maximum number of tries (`max_tries`)
-   simulation variables: the simulation number (`sim`), the seed used
    (`seed`)
-   optimisation variables: the projection basis in a matrix format
    (`basis`), the index value (`index_val`), a description of the
    status - one of "initiation", "current_best", and "jellyfish_update"
    (`info`), current iteration ID (`tries`), current jelly ID (`loop`),
    and the time taken to find the optimum (`time`).

The basis column records every basis *visited* by the jellyfish
optimiser prior to comparing with the current basis. In each iteration,
if the index value of a visited basis is smaller than that of the
current one, the jellyfish optimiser will retain the current basis for
the next iteration, while still documenting the visited basis.

Numerical information to compute:

-   angular distance between the projection basis and the theoretical
    best basis,
-   the proportion of simulation that found the optimal basis,
-   the proportion of jellies, within each simulation, that found the
    optimal basis,

Visualisation to inspect:

-   inspect the basis visited by each jellyfish in the reduced PCA space,
-   inspect the final 2D projections reached by each jellyfish,
-   plot the index value against the angular distance between the
    projection basis and the theoretical best basis

Plotting the basis in the space and the projected data can help to
understand 1) whether each simulation finds the same optimum or some
simulations find local optima; and 2) whether the index function used
can detect the structure in the data and the projection contains the
structure of interest.

The visualisations above can be faceted by the projection pursuit
variables and jellyfish optimiser parameters to compare the performance
of different indexes to detect the same structure and how the jellyfish
optimiser parameters affect the optimisation process.

[example plots]

```{r}
set.seed(123456)
sin1000 <- spinebil::sinData(6, 1000) %>% scale() %>% as_tibble()
colnames(sin1000) <- paste0("V", 1:6)

sine_dcor2d_5050 <- sim_sine_6d_dcor2d |> 
  filter(n_jellies == 50, max_tries == 50)
best_sim <-  sine_dcor2d_5050|> 
  group_by(sim) |> 
  filter(index_val == max(index_val)) |> 
  filter(row_number() == 1) 
proj_dt <- map_dfr(best_sim |> pull(basis), 
              ~as_tibble(as.matrix(sin1000) %*% .x), .id = "sim")

p1 <- proj_dt |>
  ggplot(aes(x = V1, y = V2)) + 
  geom_point(size = 0.1) + 
  xlim(-4, 4) + ylim(-4, 4) +
  facet_wrap(~as.numeric(sim), ncol = 10) + 
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())
```

```{r}
#| fig.height: 5 
#| fig.width: 10
#| fig.cap: sdfjsflk
p1 
```

```{r}
theo_best <- matrix(c(rep(0, 8), 1, 0, 0, 1), nrow = 6, byrow = TRUE)
angle_dcor2d <- sine_dcor2d_5050 |>
    rowwise() |>
    mutate(angle = tourr::proj_dist(theo_best, basis))

p2 <- angle_dcor2d |>
  ggplot(aes(x = angle, y = index_val)) +
  geom_point(size = 0.1) +
  theme(aspect.ratio = 1)

### 2024-04-19 optim.tex
outline of the section visualisation of jellyfish optimiser

\begin{keyword}
    projection pursuit \sep optimization \sep jellyfish
optimiser \sep data visualisation \sep 
    high-dimensional data
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

\begin{tcolorbox}[enhanced jigsaw, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, opacitybacktitle=0.6, leftrule=.75mm, toptitle=1mm, coltitle=black, breakable, arc=.35mm, left=2mm, rightrule=.15mm]

@@

\hypertarget{sec-simulation}{%
\section{Application {[}Di and Sherry{]}}\label{sec-simulation}}

@@

\begin{figure}

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

\end{figure}

\begin{figure}

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-3-1.pdf}

>>>>

\begin{keyword}
    projection pursuit \sep optimization \sep jellyfish
optimiser \sep data visualisation \sep 
    high-dimensional data
\end{keyword}
\end{frontmatter}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, title={Algorithm: Jellyfish Optimizer Pseudo Code}, leftrule=.75mm, breakable, titlerule=0mm, opacityback=0, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, arc=.35mm, toprule=.15mm, opacitybacktitle=0.6, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, bottomrule=.15mm, left=2mm, coltitle=black]

@@

\section{Visualisation of jellyfish
optimiser}\label{visualisation-of-jellyfish-optimiser}

Information of the jellyfish optimiser is available in a tabular format
and below is an example data collected from finding the sine-wave
structure in 6D data using a distance correlation index
(\texttt{docr2d\_2}):

\begin{verbatim}
Rows: 275,000
Columns: 13
$ idx_f     <chr> "dcor2d_2", "dcor2d_2", "dcor2d_2", "dcor2d_2", "dcor2d_2", ~
$ d         <dbl> 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, ~
$ n_jellies <dbl> 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, ~
$ max_tries <dbl> 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, ~
$ sim       <int> 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, ~
$ seed      <int> 3462, 3462, 3462, 3462, 3462, 3462, 3462, 3462, 3462, 3462, ~
$ basis     <list> <<matrix[6 x 2]>>, <<matrix[6 x 2]>>, <<matrix[6 x 2]>>, <<~
$ index_val <dbl> 0.0247212373, 0.0033938502, 0.0463398915, 0.0486230801, -0.0~
$ info      <chr> "initiation", "initiation", "initiation", "initiation", "ini~
$ method    <chr> "search_jellyfish", "search_jellyfish", "search_jellyfish", ~
$ tries     <dbl> 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, ~
$ loop      <dbl> 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 1~
$ time      <drtn> 35.46031 secs, 35.46031 secs, 35.46031 secs, 35.46031 secs,~
\end{verbatim}

Information recorded can be categorised into the following categories:

\begin{itemize}
\tightlist
\item
  projection pursuit variables: the index function used
  (\texttt{idx\_f}), the data dimension (\texttt{d})
\item
  jellyfish optimiser parameters: the number of jellies
  (\texttt{n\_jellies}), the maximum number of tries
  (\texttt{max\_tries})
\item
  simulation variables: the simulation number (\texttt{sim}), the seed
  used (\texttt{seed})
\item
  optimisation variables: the projection basis in a matrix format
  (\texttt{basis}), the index value (\texttt{index\_val}), a description
  of the status - one of ``initiation'', ``current\_best'', and
  ``jellyfish\_update'' (\texttt{info}), current iteration ID
  (\texttt{tries}), current jelly ID (\texttt{loop}), and the time taken
  to find the optimum (\texttt{time}).
\end{itemize}

The basis column records every basis \emph{visited} by the jellyfish
optimiser prior to comparing with the current basis. In each iteration,
if the index value of a visited basis is smaller than that of the
current one, the jellyfish optimiser will retain the current basis for
the next iteration, while still documenting the visited basis.

Numerical information to compute:

\begin{itemize}
\tightlist
\item
  angular distance between the projection basis and the theoretical best
  basis,
\item
  the proportion of simulation that found the optimal basis,
\item
  the proportion of jellies, within each simulation, that found the
  optimal basis,
\end{itemize}

Visualisation to inspect:

\begin{itemize}
\tightlist
\item
  inspect the basis visited by each jellyfish in the reduced PCA space,
\item
  inspect the final 2D projections reached by each jellyfish,
\item
  plot the index value against the angular distance between the
  projection basis and the theoretical best basis
\end{itemize}

Plotting the basis in the space and the projected data can help to
understand 1) whether each simulation finds the same optimum or some
simulations find local optima; and 2) whether the index function used
can detect the structure in the data and the projection contains the
structure of interest.

The visualisations above can be faceted by the projection pursuit
variables and jellyfish optimiser parameters to compare the performance
of different indexes to detect the same structure and how the jellyfish
optimiser parameters affect the optimisation process.

{[}example plots{]}

\begin{figure}[H]

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-4-1.pdf}

\caption{sdfjsflk}

\end{figure}%

\section{Application {[}Di and Sherry{]}}\label{sec-simulation}

@@

\begin{figure}[H]

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-7-1.pdf}

@@

\end{figure}%

\begin{figure}[H]

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-8-1.pdf}
